The Development of a Collaborative Tool to Teach Debugging

### 2020-12-06 paper/paper.tex
Paper before final read through

\begin{itemize}
\item Keith O'Hara, my senior project advisor, for his constant
  thoughtfulness about my project, his advice about the future, and
  for giving me the oppurtunity to tutor for his classes.
\item Mar Hicks and Kyle Hale, professors whose classes at IIT changed
  my perspective on computer science.
\item All my professors at Bard, particuarly Olga Voronina, Wakako
  Suzuki, and Nate Shockey, whose encouragement of my interests
  outside of computer science has been instrumental in continuing my
  passion for it.
\item My family and friends, for their constant love and support.
\end{itemize}

@@

Programmers spend large amounts of time debugging.  They often use
debuggers to assist them in this process of diagnosing sometimes
inexplicable problems with their code.  Though debuggers are primarily
used to hunt down hiden mistakes, they are also powerful tools to
understand program execution.
\par

@@

There is an unfortunate lack of research specifically into the
efficacy of teaching debugging for computer science students, despite
a recent rise in the inclusion of debugging in ``computational
thinking'' curriculums\cite{10.1145/3361721.3361724}.  These
curriculums attempt to teach skills in computer science classes that
translate into other subject areas: the UK's computer science
curriculum considers debugging an essential ``transferable
skill''\cite{10.1145/2602484}.\par

@@

Research into the value of collaborative programming is overwhelmingly
positive.  McDowell et. al. found that not only does collaborative
programming significantly boost student confidence and the retention
of students in computer science majors, but that it demonstrably
improves student's work \cite{10.1145/1145287.1145293}.  These
benefits of confidence and correctness are present when paired and
non-paired students are given identical assignments
\cite{10.1145/1026487.1008043}, further indicating that the simple act
of collaboration definitively benefits computer science students.
\par

It seems that the benefits in confidence that result from teaching
debugging should be magnified by teaching debugging collaboratively.
There are similarities between the introductory nature of teaching
fundemental debugging skills and the nature of teaching fundemental
programming skills covered in the classes studied in
\cite{10.1145/1026487.1008043} and \cite{10.1145/1145287.1145293}.
Introductory programming classes typically use a specific programming
language in order to introduce widely applicable programming
principles.  By using a specific collaborative platform to introduce
debugging principles, students may realize the same benefits in both
confidence \textit{and} correctness that they do from pair programming
in introductory comptuer science classes.
\par

@@

\section{Requirments}

In order to provide the best platform for both teachers and students,
the collaborative debugger must fufill two key requirements:

@@

\begin{enumerate}
\item \lstinline{hash}: an example of classic pointer confusion.  The
  hashtable implementation included with the program is slightly
  bugged, where failing to call \lstinline{strncpy} results in the
  value for each key-value pair in the hash pointing to the same
  location in memory.  This is particuarly useful example for learning
  about setting watchpoints and breakpoints, as well as basic
  procedures to step through code.
\item \lstinline{smash}: a brief example of stack smashing.  A badly
  designed call to \lstinline{sscanf} results in a portion of the call
  stack being overwritten, and the process is killed as a result of
  stack smashing.  This example focuses on printing and examining
  memory locations.
\item \lstinline{memoization}: a program that demonstrates the
  difference between a recursive, iterative, and memoized recursive
  implimentation of the classic nth fibonacci number function.  This
  example focuses on examining the stack under conditions produced by
  different types of functions.
\end{enumerate}

@@

Students using all three examples would benefit greatly from
visualization tools to help understand pointer locations, the process
space, and data structures.  As discussed in \ref{vistools}, adding
vizualization tools to the collaborative debugger should be relatively
easy.
\par

@@

Notice that the debugging window in the example shown above is called
``rr''.  rr is an enhancment to gdb that is foundational in the
collaborative debugger's ability to deterministically record and
replay program execution. rr and its role in the collaborative
debugger will be covered extensively in (\ref{rrpods}).  Users
interact with rr through the collaborative debugger's frontend.  Care
has been taken to make the frontend as representative of rr as
possible so that users can learn to use rr/gdb by using the
collaborative debugger.

@@

The frontend of the collaborative debugger must handle constant
changes in state as users participate in the process outlined above.
The frontend uses the following tools to manage state and display
output:

@@

React makes it easy to create simple webapps with no build
configuration necessary through the use of the
\lstinline{create-react-app} command.  This creates a structure and
toolchain for a basic one-page application---suitable for the
collaborative debugger.  Once this structure was created and any
superfluous files/code were removed, it's wasn't necessary for the
purposes of the collaborative debugger to directly edit any files
except two: \lstinline{App.js} and \lstinline{App.css}.  The
pre-defined build process takes care of transforming the React code
defined in \lstinline{App.js} and the styles defined in
\lstinline{App.css} into portable Javascript, HTML, and CSS.
\par

@@

In debugging the \lstinline{memoization} example program, users will
want to display the call stack from within each implementation of the
nth fibonacci number function.  To stop in \lstinline{fib_rec}, the
recursive implementation, one user will want to set a breakpoint at
line 52 in the source file.  To do this they will type the command
\lstinline{break 52} and hit enter.  The command will be sent to the
backend, and all users in the same debug session (17734 in Figure
\ref{frontend:rrterm}) will see the debuggers response:\\
\lstinline{Breakpoint 1 at 0x561c8fede303: file memo.c, line 52.}\\
Users who did not send the command will see it displayed along with
the response.

@@

\lstinline{name}, \lstinline{linum}, and \lstinline{code} store
information about the current source file being debugged.
\lstinline{command} is the most important field for examing how
commands are sent to the backend, as it holds the current command.
\textit{Every time} the user types a character, \lstinline{command} is
updated to reflect the current state of the Xterm.js component.  The
code to handle keystrokes is shown below:

@@

Whenever the user types a printable key, the key is written to the
terminal and \lstinline{command} is updated using the
\lstinline{updateCommand} method.  If a user types \lstinline{brk},
\lstinline{updateCommand} would be called three times, once for each
keystroke.  Upon realizing they made a mistake when typing
\lstinline{break 52}, the user would likely delete a character.  As
long as the prompt (\lstinline{(rr)} followed by a space) would not be
overwritten, the terminal is updated and \lstinline{updateCommand} is
called again.  When the user hits the enter key the command is finally
sent to the backend using \lstinline{sendCommand}.  To mimick the
behavior of a local rr debug session and to prevent garbled input,
writing to the terminal is disabled until a response is recieved.

@@

All users connected to the same debug session will recieive responses
to the commands sent by one user.  This is how collaboration is
achieved in the collaborative debugger.  The JSON response object to
\lstinline{break 52} recieved by all users is shown below:

@@

Care has been taken to design the frontend in a modular fashion so
that it can be easily extended in the future.  By sending additional
data about program state in the response and creating new componenets,
visualization tools could be easily added.  If further authentication
features or the ability to edit source code are desired in the future,
relevant components can be updated without the need to change the
entire application.  Since the collaborative debugger depends heavily
on interactions between it's frontend and backend, similar care has
been taken in designing the backend, which will be described in the
following sections.

@@

Kubernetes is the defacto standard in container orchestration
software.  It provides a layer of abstraction on top of normal
containers, like those created by Docker.  By bundling one or more
closely linked containers into a ``Pod'', Kubernetes is able to manage
deployment and re-deployment of applications running inside
containers.  It is trivial to create new Pods, or to create multiple
Pods running the same application as needed within a Kubernetes
cluster \cite{k8s}.  The speed at which even relatively large Pods can
be created and the inherent security provided by containerization
drove the decision to create a new Pod on the fly for each debugging
instance in the collaborative debugger.  This allows the debugger to
provide a similar level of convinience to exisiting collaborative
tools, such repl.it (\ref{exisitingcollab}).
\par

@@

The collaborative debugger consists of a distributed Kubernetes
backend and frontend React webapp.  Kubernetes was chosen for the
backend primarally so that a Pod could be created dynamically for each
debugging session.

@@

\begin{enumerate}
\item The \textit{Database ClusterIP}: a ClusterIP, which exposes the
  Database Pod only inside the cluster.  The only component that makes
  use of this ClusterIP is the API server, which uses it primarily to
  communicate information about users and RR Debug Sessions with the
  database.
\item The \textit{RR Message Load Balancer}: a Load Balancer, which
  exposes the RR Message Server Pods to the outside world.  Using
  Socket.IO, clients send commands to and recieve responses from
  individual RR Debug Session Pods through the RR Message Server Pods.
\item The \textit{Frontend Load Balancer}: another Load Balancer,
  which exposes the Frontend/API Server Pods to the outside world.
  Clients request the frontend webapp and send api requests/recieve
  api responses through this Load Balancer.
\end{enumerate}

@@

\begin{enumerate}
\item The base image is defined. The RR Message Server and RR Debug
  Session images are based on the latest Ubuntu image.  This is
  particuarly necessary for the RR Debug Session image, as rr's
  low-level nature necessitates somewhat frequent updates as changes
  are made to the Linux kernel.  The Frontend/API Server image is
  based on the latest Node image, and the Database image is based on
  the latest MongoDB image.
\item Second, any necessary packages are installed.  For the RR Debug
  Pod image, rr is compiled from source and installed.
\item A non-root user is created if necessary.
\item Program files are copied over and a file structure is created.
  Packages that don't rely on the base images builtin package manager
  are installed at this time.  In the example above, these include
  Python packages.
\item A startup command is defined.
\end{enumerate}

@@

Services are the first components of the collaborative debugger
created after the cluster is initiallized.  The three Services used by
the debugger are all statically created.  Like statically created
Pods, schemas that define Services are applied manually using the
\lstinline{kubectl apply} command.

@@

At the heart of the collaborative debugger is the RR Debug Session.
Every time a user wants to debug a new program execution, an RR Debug
Session Pod is dynamically created by the API server.  The new Pod is
assigned a unique five digit identification number when it is created.
This five digit number, is used as the channel for the RR Debug
Session, seperating its communication from other RR Debug Session
Pods.  Clients, RR Message Server Pods, and the RR Debug Session Pod
connect through the RR Message Load Balancer.  Clients and RR Debug
Sessions send messages using the channel assigned to the pod at
creation time.  A diagram of one full communication cycle between two
clients and an RR Debug Session Pod is outlined in the diagram below:

@@

\begin{enumerate}
\item Client A, already having connected to the channel that
  corresponds to it's current RR Debug Session (channel 12345 in the
  example), sends an rr command.
\item The RR Message Server recieves the command and emits a message
  that RR Debug Session Pods are equiped to recieve on the same
  channel.  In practice, since only one RR Debug Session Pod is ever
  on a channel, this equates to emmiting a message to the Pod
  directly.
\item The RR Debug Session Pod recieves the message, and passes the
  command to its instance of RRInterface.  When it recieves a
  response, it emits the output from rr along with other debugging
  information and the channel.
\item The RR Message Server recieves the response, and emits a
  response message that clients are equipped to recieve on the
  corresponding channel.
\item Clients A and B are connected to the corresponding channel, so
  they both recieve the response.
\end{enumerate}

@@

The collaborative debugger needs an effecient way for clients to
communicate with running RR Debug Sessions. The Socket.IO library was
selected to construct this efficient communication pathway.

@@

To speed communication, the collaborative debugger uses WebSockets to
directly connect web clients and the Pods running rr.  Socket.IO is a
library that extends WebSockets.  It provides backup incase a
WebSocket connection cannot be established, enables automatic
reconnection and disconnection detection, and adds support for
namespaces \cite{socketio}.  The collaborative debugger uses the
standard JavaScript implementation of Socket.IO on the client side.
Messages are passed through a server to individual debugging Pods,
both of which use the Python implementation of Socket.IO,
python-socket.io \cite{python_socketio}.

@@

The two message processing functions, \lstinline{on_rr_command} and
\lstinline{on_rr_response}, are equally simple.  When the RR Message
Server recieves an \lstinline{'rr_command'} message, it passes the
message data along to the corresponding RR Debug Session Pod by
emitting an \lstinline{'rr_command'} namespaced to the correct channel
with \lstinline{room=data['channel']}.  Since the socket.io client
running in RR Message Server Pods has an event handler defined for
\lstinline{'rr_command'}s but frontend clients do not, only RR Message
Server Pods recieve the command.  The same process happens in reverse
for \lstinline{'rr_response'}s, with only the frontend socket.io
clients having a handler defined for \lstinline{'rr_response'}.

@@

\paragraph{Mozilla's rr}\label{rr}

rr is ``a lightweight tool for recording, replaying and debugging
execution of applications''\cite{rr-repo}. rr allows a programmer to
record the execution of a program on any compatible machine and replay
the execution later.  This enhances GDB's ability to ``time-travel''
when debugging, using commands such as \lstinline{reverse-continue}
and \lstinline{reverse-stepi}\cite{gdbman} to step backwards and
forwards through a program's execution.  Through a novel encapsulation
of the execution space, rr is able to deterministically record and
replay the execution of syscalls and other process behavior that
differs run-to-run.  This is invaluable when trying to debug behavior
that is not entirely dependent on the code being debugged.  A typical
workflow in rr consists of recording an inexplicable error, replaying
execution to find the area in which the error occurs, and then
narrowing in on the bug not by re-running the entire program, but by
progressing back and forth through execution in the problem area.
\par

Luckily, the speed and size benefits of rr lend themselves well to
non-local execution.  In conjunction with Kubernetes, it takes a few
seconds to create a new container running rr and connect to web
clients.

@@

In order to ``support the development of systems which use the
debugger as just one small component of a larger system'', gdb
provides a machine-oriented interface called gdb/mi \cite{gdbman}. rr
supports interaction through gdb/mi, and using the interface was a
natural choice for the collaborative debugger.  In addition to being
far easier to interact with from within a program, the structured,
machine-friendly output of gdb/mi lends itself in particular to future
development of visualization aids in the collaborative debugger.
\par
To parse rr output into Python dictionaries and to easily control rr
as a subprocess, pygdbmi \cite{pygdbmi} is used in each debugging Pod.
pygdbmi's abstraction simplifies programatically controlling rr.  A
Pod can receive a command from the client, pass it to rr, and respond
without having to deal with parsing gdb/mi output or with directly
managing the rr process.

@@

The process for building RR Debug Session example images is differs
slightly from other images used by the collaborative debugger.
Currently, three example RR Debug Session images are availiable for
use.  Example container images are based on the primary RR Debug
Session image, RR Translation.  The build process for this image,
based on the lasted official Ubuntu container image, installs rr,
python, and all necessary packages as well as the app that will run on
the final image, \lstinline{rrtranslation.py}.

@@

After joining the appropriate channel, \lstinline{sio} waits to
recieve an \lstinline{'rr_command'}.  The first part of
\lstinline{sio}'s \lstinline{'rr_command'} event handler is shown below:

@@

The handler first builds part of the response body, passing back data
about the channel and originator of the message that will be important
for the RR Message Server and client later.  Next, it calls the
function necessary to pass a message to rr and recieve a response,
\lstinline{rri.write()}.
\par

@@

Next, write appends the command to the \lstinline{RRInterface}'s
\lstinline{timline} instance variable.  \lstinline{timeline} is a list
of all commands the debug session has executed.  Though unused at the
moment, it can serve in the future to impliment a shared history
between all clients and to facilitate the saving of debug sessions
in-progress.  By issuing all commands in \lstinline{timeline} to a new
instance of \lstinline{RRInterface} with the same recording, it is
possible to restore a debug session to an identical previous state.
To save a debug session, the recording and \lstinline{timeline} can be
stored in a database, and be used to initialize a new Pod with
identical state to a previous RR Debug Session Pod.  This appears to
clients as a seemless restoration of a previous Debug Session.
\par

Since the frontend currently consists of a terminal interface to rr
and a view of the current source code, write does not need to return
any information from rr other than console output.  Future versions of
the collaborative debugger that support visualization tools could use
gdbmi commands such as \lstinline{-stack-list-frames} to retrieve
information to be used by frontend visualization tools.
\lstinline{console_output} extracts user-relevant output from the
dictionary returned by \lstinline{GDBController}'s \lstinline{write}
method.  The \lstinline{get_full_rr_response} method ensures that all
relavant output has been recieved from the rr subprocess before
returning.  Write reduces the lengthy dictionary that would be
produced by a command as simple as \lstinline{break main} into a few
lines of user-relevant output.
\par

After the \lstinline{'rr_command'} event handler has recieved a
response from \lstinline{rri}, it calls \lstinline{RRInterface}'s
\lstinline{source} function to retrieve information about the current
source file being debugged.  Most often the source file has not
changed, and the only relevant piece of information that needs to be
passed back to the client is the new line number in the source file.
If the source file has changed, the handler reads its contents,
updates \lstinline{current_file}, and emits the line number, file
name, and contents.  If an error occurs at any point in the process, a
detailed trace is emmitted for debugging purposes.  The trace should
be ommited in production.
\par

@@

Most of the complexity in the RR Translation program stems from
parsing rr output.  The flow of data in the program is quite simple: a
command is recieved, it is passed to rr, rr's response is processed,
and a response message is emmited.  This flow would remain unchanged
even with the addition of functions to retrieve information for data
visualization.  Since the process of turning rr responses into a
terminal interface takes place in the frontend, rather than the RR
Translation program providing some sort of REPL over WebSockets,
updates can be made to the frontend webapp without requiring backend
changes.  The whole process is also quite snappy, with only a slight
delay from the client's perspective compared to running a debugger
natively.
\par

Effort is taken in the design of the RR Debug Session and RR Message
Server to ensure students recieve a pleasant and near-identical
experience to debugging locally in the future.  The system of
inheritance used to create example debug session pods is designed to
be as simple as possible, though a more fully featured frontend could
simplify this process further in the future.

@@

Since the entire webapp is one page, the process for serving it is
completely standard.  The server recieves a request for the index page
of the website, and returns the compiled React webapp that makes up
the homepage.  The process for API requests is more involved, and an
example is outlined below:

@@

The API server is implimented using Flask.  The server program
consists of handlers for the various API routes and instances of two
classes which communicate with the database/Kubernetes API:
\lstinline{TranslationPodManager} and \lstinline{UserManager}.  The
full structure of the API Server program, \lstinline{api.py} is shown
below:

@@

All API routes only support POST requests.  The route handler first
extracts the relavant information from the POST request body, in this
case \lstinline{name} and \lstinline{program}.  \lstinline{name}
always corresponds the user who is making the request's username, and
\lstinline{program} corresponds to the name of the example RR Debug
Session image to be used.
\par

@@

>>>>

\begin{itemize}
\item Keith O'Hara, my senior project advisor, for his constant
  thoughtfulness about my project, his advice about the future, and
  for giving me the opportunity to tutor for his classes.
\item Mar Hicks and Kyle Hale, professors whose classes at IIT changed
  my perspective on computer science.
\item All my professors at Bard, particularly Olga Voronina, Wakako
  Suzuki, and Nate Shockey, whose encouragement of my interests
  outside of computer science has been instrumental in continuing my
  passion for it.
\item My family and friends, for their constant love and support.
\end{itemize}

@@

Programmers spend large amounts of time debugging---the process of
searching for and correcting errors in code that are not immediately
apparent.  They often use debuggers---tools which help programmers to
inspect a running program---to assist them in this process of
diagnosing sometimes inexplicable problems with their code.  Though
debuggers are primarily used to hunt down hidden mistakes, they are
also powerful tools to understand program execution.
\par

@@

There is an unfortunate lack of research specifically into the
efficacy of teaching debugging for computer science students, despite
a recent rise in the inclusion of debugging in ``computational
thinking'' curricula\cite{10.1145/3361721.3361724}.  These
curricula attempt to teach skills in computer science classes that
translate into other subject areas: the UK's computer science
curriculum considers debugging an essential ``transferable
skill''\cite{10.1145/2602484}.\par

@@

Collaborative programming, where multiple programmers work together to
write and test code, is popular in both industry and computer science
education\cite{10.1145/1145287.1145293}. Collaborative programming
most often manifests as pair programming---two programmers working
together on the same program.  Research into the value of pair
programming is overwhelmingly positive.  McDowell et. al. found that
not only does pair programming significantly boost student confidence
and the retention of students in computer science majors, but that it
demonstrably improves student's work \cite{10.1145/1145287.1145293}.
These benefits of confidence and correctness are present when paired
and non-paired students are given identical assignments
\cite{10.1145/1026487.1008043}, further indicating that the simple act
of collaboration definitively benefits computer science students.
\par

It seems that the benefits in confidence that result from teaching
debugging should be magnified by teaching debugging collaboratively.
There are similarities between the introductory nature of teaching
fundamental debugging skills and the nature of teaching fundamental
programming skills covered in the classes studied in
\cite{10.1145/1026487.1008043} and \cite{10.1145/1145287.1145293}.
Introductory programming classes typically use a specific programming
language in order to introduce widely applicable programming
principles.  By using a specific collaborative platform to introduce
debugging principles, students may realize the same benefits in both
confidence \textit{and} correctness that they do from pair programming
in introductory computer science classes.
\par

@@

\section{Requirements}

In order to provide the best platform for both teachers and students,
the collaborative debugger must fulfill two key requirements:

@@

\begin{enumerate}
\item \lstinline{hash}: an example of classic pointer confusion.  The
  hashtable implementation included with the program is slightly
  bugged, where failing to call \lstinline{strncpy} results in the
  value for each key-value pair in the hash pointing to the same
  location in memory.  This is particularly useful example for learning
  about setting watchpoints and breakpoints, as well as basic
  procedures to step through code.
\item \lstinline{smash}: a brief example of stack smashing.  A badly
  designed call to \lstinline{sscanf} results in a portion of the call
  stack being overwritten, and the process is killed as a result of
  stack smashing.  This example focuses on printing and examining
  memory locations.
\item \lstinline{memoization}: a program that demonstrates the
  difference between a recursive, iterative, and memoized recursive
  implementation of the classic nth Fibonacci number function.  This
  example focuses on examining the stack under conditions produced by
  different types of functions.
\end{enumerate}

@@

Students using all three examples would benefit greatly from
visualization tools to help understand pointer locations, the process
space, and data structures.  As discussed in \ref{vistools}, adding
visualization tools to the collaborative debugger should be relatively
easy.
\par

@@

@@

The frontend of the collaborative debugger must handle constant
changes in state as users participate in the process outlined above.
The frontend uses React, Monaco, and Xterm.js to manage state and
display output.  It primarily displays the output from rr, an
enhancement of gdb that the collaborative debugger uses to enable
collaboration.

\subsubsection{Mozilla's rr}\label{rr}

Notice that the debugging window in the above example is called
``rr''. rr is ``a lightweight tool for recording, replaying and
debugging execution of applications''\cite{rr-repo}. rr allows a
programmer to record the execution of a program on any compatible
machine and replay the execution later.  This enhances GDB's ability
to ``time-travel'' when debugging, using commands such as
\lstinline{reverse-continue} and
\lstinline{reverse-stepi}\cite{gdbman} to step backwards and forwards
through a program's execution.  Through a novel encapsulation of the
execution space, rr is able to deterministically record and replay the
execution of syscalls and other process behavior that differs
run-to-run.  This is invaluable when trying to debug behavior that is
not entirely dependent on the code being debugged.  A typical workflow
in rr consists of recording an inexplicable error, replaying execution
to find the area in which the error occurs, and then narrowing in on
the bug not by re-running the entire program, but by progressing back
and forth through execution in the problem area.
\par

Luckily, the speed and size benefits of rr lend themselves well to
non-local execution.  In conjunction with Kubernetes (\ref{k8s}), it
takes a few seconds to create a new container (\ref{docker}) running
rr and connect to web clients.

@@

React makes it easy to create simple webapps with no build
configuration necessary through the use of the
\lstinline{create-react-app} command.  This creates a structure and
toolchain for a basic one-page application---suitable for the
collaborative debugger.  Once this structure was created and any
superfluous files/code were removed, it's wasn't necessary for the
purposes of the collaborative debugger to directly edit any files
except two: \lstinline{App.js} and \lstinline{App.css}.  The
pre-defined build process takes care of transforming the React code
defined in \lstinline{App.js} and the styles defined in
\lstinline{App.css} into portable JavaScript, HTML, and CSS.
\par

@@

In debugging the \lstinline{memoization} example program, users will
want to display the call stack from within each implementation of the
nth Fibonacci number function.  To stop in \lstinline{fib_rec}, the
recursive implementation, one user will want to set a breakpoint at
line 52 in the source file.  To do this they will type the command
\lstinline{break 52} and hit enter.  The command will be sent to the
backend, and all users in the same debug session (17734 in Figure
\ref{frontend:rrterm}) will see the debuggers response:\\
\lstinline{Breakpoint 1 at 0x561c8fede303: file memo.c, line 52.}\\
Users who did not send the command will see it displayed along with
the response.

@@

\lstinline{name}, \lstinline{linum}, and \lstinline{code} store
information about the current source file being debugged.
\lstinline{command} is the most important field for examining how
commands are sent to the backend, as it holds the current command.
\textit{Every time} the user types a character, \lstinline{command} is
updated to reflect the current state of the Xterm.js component.  The
code to handle keystrokes is shown below:

@@

Whenever the user types a printable key, the key is written to the
terminal and \lstinline{command} is updated using the
\lstinline{updateCommand} method.  If a user types \lstinline{brk},
\lstinline{updateCommand} would be called three times, once for each
keystroke.  Upon realizing they made a mistake when typing
\lstinline{break 52}, the user would likely delete a character.  As
long as the prompt (\lstinline{(rr)} followed by a space) would not be
overwritten, the terminal is updated and \lstinline{updateCommand} is
called again.  When the user hits the enter key the command is finally
sent to the backend using \lstinline{sendCommand}.  To mimic the
behavior of a local rr debug session and to prevent garbled input,
writing to the terminal is disabled until a response is received.

@@

All users connected to the same debug session will receive responses
to the commands sent by one user.  This is how collaboration is
achieved in the collaborative debugger.  The JSON response object to
\lstinline{break 52} received by all users is shown below:

@@

Care has been taken to design the frontend in a modular fashion so
that it can be easily extended in the future.  By sending additional
data about program state in the response and creating new components,
visualization tools could be easily added.  If further authentication
features or the ability to edit source code are desired in the future,
relevant components can be updated without the need to change the
entire application.  Since the collaborative debugger depends heavily
on interactions between it's frontend and backend, similar care has
been taken in designing the backend, which will be described in the
following sections.

@@

Kubernetes is the de facto standard in container orchestration
software.  It provides a layer of abstraction on top of normal
containers, like those created by Docker.  By bundling one or more
closely linked containers into a ``Pod'', Kubernetes is able to manage
deployment and re-deployment of applications running inside
containers.  It is trivial to create new Pods, or to create multiple
Pods running the same application as needed within a Kubernetes
cluster \cite{k8s}.  The speed at which even relatively large Pods can
be created and the inherent security provided by containerization
drove the decision to create a new Pod on the fly for each debugging
instance in the collaborative debugger.  This allows the debugger to
provide a similar level of convenience to existing collaborative
tools, such Replit (\ref{exisitingcollab}).
\par

@@

The collaborative debugger consists of a distributed Kubernetes
backend and frontend React webapp.  Kubernetes was chosen for the
backend primarily so that a Pod could be created dynamically for each
debugging session.

@@

\begin{enumerate}
\item The \textit{Database ClusterIP}: a ClusterIP, which exposes the
  Database Pod only inside the cluster.  The only component that makes
  use of this ClusterIP is the API server, which uses it primarily to
  communicate information about users and RR Debug Sessions with the
  database.
\item The \textit{RR Message Load Balancer}: a Load Balancer, which
  exposes the RR Message Server Pods to the outside world.  Using
  Socket.IO, clients send commands to and receive responses from
  individual RR Debug Session Pods through the RR Message Server Pods.
\item The \textit{Frontend Load Balancer}: another Load Balancer,
  which exposes the Frontend/API Server Pods to the outside world.
  Clients request the frontend webapp and send API requests/receive
  API responses through this Load Balancer.
\end{enumerate}

@@

\begin{enumerate}
\item The base image is defined. The RR Message Server and RR Debug
  Session images are based on the latest Ubuntu image.  This is
  particularly necessary for the RR Debug Session image, as rr's
  low-level nature necessitates somewhat frequent updates as changes
  are made to the Linux kernel.  The Frontend/API Server image is
  based on the latest Node image, and the Database image is based on
  the latest MongoDB image.
\item Second, any necessary packages are installed.  For the RR Debug
  Pod image, rr is compiled from source and installed.
\item A non-root user is created if necessary.
\item Program files are copied over and a file structure is created.
  Packages that don't rely on the base images builtin package manager
  are installed at this time.  In the example above, these include
  Python packages.
\item A startup command is defined.
\end{enumerate}

@@

Services are the first components of the collaborative debugger
created after the cluster is initialized.  The three Services used by
the debugger are all statically created.  Like statically created
Pods, schemas that define Services are applied manually using the
\lstinline{kubectl apply} command.

@@

At the heart of the collaborative debugger is the RR Debug Session.
Every time a user wants to debug a new program execution, an RR Debug
Session Pod is dynamically created by the API server.  The new Pod is
assigned a unique five digit identification number when it is created.
This five digit number, is used as the channel for the RR Debug
Session, separating its communication from other RR Debug Session
Pods.  Clients, RR Message Server Pods, and the RR Debug Session Pod
connect through the RR Message Load Balancer.  Clients and RR Debug
Sessions send messages using the channel assigned to the pod at
creation time.  A diagram of one full communication cycle between two
clients and an RR Debug Session Pod is outlined in the diagram below:

@@

\begin{enumerate}
\item Client A, already having connected to the channel that
  corresponds to it's current RR Debug Session (channel 12345 in the
  example), sends an rr command.
\item The RR Message Server receives the command and emits a message
  that RR Debug Session Pods are equipped to receive on the same
  channel.  In practice, since only one RR Debug Session Pod is ever
  on a channel, this equates to emitting a message to the Pod
  directly.
\item The RR Debug Session Pod receives the message, and passes the
  command to its instance of RRInterface.  When it receives a
  response, it emits the output from rr along with other debugging
  information and the channel.
\item The RR Message Server receives the response, and emits a
  response message that clients are equipped to receive on the
  corresponding channel.
\item Clients A and B are connected to the corresponding channel, so
  they both receive the response.
\end{enumerate}

@@

The collaborative debugger needs an efficient way for clients to
communicate with running RR Debug Sessions. The Socket.IO library was
selected to construct this efficient communication pathway.

@@

To speed communication, the collaborative debugger uses WebSockets to
directly connect web clients and the Pods running rr.  Socket.IO is a
library that extends WebSockets.  It provides backup in case a
WebSocket connection cannot be established, enables automatic
reconnection and disconnection detection, and adds support for
namespaces \cite{socketio}.  The collaborative debugger uses the
standard JavaScript implementation of Socket.IO on the client side.
Messages are passed through a server to individual debugging Pods,
both of which use the Python implementation of Socket.IO,
python-socket.io \cite{python_socketio}.

@@

The two message processing functions, \lstinline{on_rr_command} and
\lstinline{on_rr_response}, are equally simple.  When the RR Message
Server receives an \lstinline{'rr_command'} message, it passes the
message data along to the corresponding RR Debug Session Pod by
emitting an \lstinline{'rr_command'} namespaced to the correct channel
with \lstinline{room=data['channel']}.  Since the socket.io client
running in RR Message Server Pods has an event handler defined for
\lstinline{'rr_command'}s but frontend clients do not, only RR Message
Server Pods receive the command.  The same process happens in reverse
for \lstinline{'rr_response'}s, with only the frontend socket.io
clients having a handler defined for \lstinline{'rr_response'}.

@@

@@

In order to ``support the development of systems which use the
debugger as just one small component of a larger system'', gdb
provides a machine-oriented interface called gdb/mi \cite{gdbman}. rr
supports interaction through gdb/mi, and using the interface was a
natural choice for the collaborative debugger.  In addition to being
far easier to interact with from within a program, the structured,
machine-friendly output of gdb/mi lends itself in particular to future
development of visualization aids in the collaborative debugger.
\par
To parse rr output into Python dictionaries and to easily control rr
as a subprocess, pygdbmi \cite{pygdbmi} is used in each debugging Pod.
pygdbmi's abstraction simplifies programmatically controlling rr.  A
Pod can receive a command from the client, pass it to rr, and respond
without having to deal with parsing gdb/mi output or with directly
managing the rr process.

@@

The process for building RR Debug Session example images is differs
slightly from other images used by the collaborative debugger.
Currently, three example RR Debug Session images are available for
use.  Example container images are based on the primary RR Debug
Session image, RR Translation.  The build process for this image,
based on the lasted official Ubuntu container image, installs rr,
python, and all necessary packages as well as the app that will run on
the final image, \lstinline{rrtranslation.py}.

@@

After joining the appropriate channel, \lstinline{sio} waits to
receive an \lstinline{'rr_command'}.  The first part of
\lstinline{sio}'s \lstinline{'rr_command'} event handler is shown below:

@@

The handler first builds part of the response body, passing back data
about the channel and originator of the message that will be important
for the RR Message Server and client later.  Next, it calls the
function necessary to pass a message to rr and receive a response,
\lstinline{rri.write()}.
\par

@@

Next, write appends the command to the \lstinline{RRInterface}'s
\lstinline{timline} instance variable.  \lstinline{timeline} is a list
of all commands the debug session has executed.  Though unused at the
moment, it can serve in the future to implement a shared history
between all clients and to facilitate the saving of debug sessions
in-progress.  By issuing all commands in \lstinline{timeline} to a new
instance of \lstinline{RRInterface} with the same recording, it is
possible to restore a debug session to an identical previous state.
To save a debug session, the recording and \lstinline{timeline} can be
stored in a database, and be used to initialize a new Pod with
identical state to a previous RR Debug Session Pod.  This appears to
clients as a seamless restoration of a previous Debug Session.
\par

Since the frontend currently consists of a terminal interface to rr
and a view of the current source code, write does not need to return
any information from rr other than console output.  Future versions of
the collaborative debugger that support visualization tools could use
gdb/mi commands such as \lstinline{-stack-list-frames} to retrieve
information to be used by frontend visualization tools.
\lstinline{console_output} extracts user-relevant output from the
dictionary returned by \lstinline{GDBController}'s \lstinline{write}
method.  The \lstinline{get_full_rr_response} method ensures that all
relevant output has been received from the rr subprocess before
returning.  Write reduces the lengthy dictionary that would be
produced by a command as simple as \lstinline{break main} into a few
lines of user-relevant output.
\par

After the \lstinline{'rr_command'} event handler has received a
response from \lstinline{rri}, it calls \lstinline{RRInterface}'s
\lstinline{source} function to retrieve information about the current
source file being debugged.  Most often the source file has not
changed, and the only relevant piece of information that needs to be
passed back to the client is the new line number in the source file.
If the source file has changed, the handler reads its contents,
updates \lstinline{current_file}, and emits the line number, file
name, and contents.  If an error occurs at any point in the process, a
detailed trace is emitted for debugging purposes.  The trace should
be omitted in production.
\par

@@

Most of the complexity in the RR Translation program stems from
parsing rr output.  The flow of data in the program is quite simple: a
command is received, it is passed to rr, rr's response is processed,
and a response message is emitted.  This flow would remain unchanged
even with the addition of functions to retrieve information for data
visualization.  Since the process of turning rr responses into a
terminal interface takes place in the frontend, rather than the RR
Translation program providing some sort of REPL over WebSockets,
updates can be made to the frontend webapp without requiring backend
changes.  The whole process is also quite snappy, with only a slight
delay from the client's perspective compared to running a debugger
natively.
\par

Effort is taken in the design of the RR Debug Session and RR Message
Server to ensure students receive a pleasant and near-identical
experience to debugging locally in the future.  The system of
inheritance used to create example debug session pods is designed to
be as simple as possible, though a more fully featured frontend could
simplify this process further in the future.

@@

Since the entire webapp is one page, the process for serving it is
completely standard.  The server receives a request for the index page
of the website, and returns the compiled React webapp that makes up
the homepage.  The process for API requests is more involved, and an
example is outlined below:

@@

The API server is implemented using Flask.  The server program
consists of handlers for the various API routes and instances of two
classes which communicate with the database/Kubernetes API:
\lstinline{TranslationPodManager} and \lstinline{UserManager}.  The
full structure of the API Server program, \lstinline{api.py} is shown
below:

@@

All API routes only support POST requests.  The route handler first
extracts the relevant information from the POST request body, in this
case \lstinline{name} and \lstinline{program}.  \lstinline{name}
always corresponds the user who is making the request's username, and
\lstinline{program} corresponds to the name of the example RR Debug
Session image to be used.
\par

@@

\subsection{Current Functionality}
